---
title: "Linear models"
author: "Léo Belzile"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    lib_dir: "libs"
    chakra: "libs/remark-latest.min.js"
    css: ["default", "css/ath-slides.css", "css/ath-inferno-fonts.css", "css/animate.css"]
    seal: false
    anchor_sections: false
    nature:
      highlightStyle: github
      highlightLines: false
      countIncrementalSlides: false
      ratio: "16:9"
      navigation:
        scroll: false
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, 
                      message = FALSE, 
                      fig.retina = 3, 
                      fig.align = "center",
                      fig.width = 10,
                      fig.asp = 0.618,
                      out.width = "70%")
```
```{r packages-data, echo = FALSE, include=FALSE}
library(knitr)
options(knitr.kable.NA = '')
options(tidyverse.quiet = TRUE)
options(knitr.table.format = "html")
library(tidyverse)
library(patchwork)
```
```{r xaringanExtra, echo=FALSE}
xaringanExtra::use_xaringan_extra(c("tile_view","freezeframe","panelset","clipboard","broadcast"))
```

class: center middle main-title section-title-1

# Linear models

.class-info[

**Session 8**

.light[MATH 80667A: Experimental Design and Statistical Methods <br>for Quantitative Research in Management <br>
HEC Montréal
]

]

---

name: outline
class: title title-inv-1

# Outline
--


.box-4.medium.sp-after-half[Model formulation]

--

.box-5.medium.sp-after-half[Sum of square decompositions for unbalanced designs]

---

.box-6.medium.sp-after-half[Analysis of trend]
--

.box-7.medium.sp-after-half[Diagnostic plots]

---
layout: false
name: model-formulation
class: center middle section-title section-title-4 animated fadeIn

# Linear model

---


class: title title-4

# IJLR: It's just a linear regression...

All models we covered so fall ($t$-tests, one- and two-way ANOVA, factorial designs, latin squares) are all very special
instances of the linear regression model.

The latter says that the expected 


