---
title: "Linear models"
author: "Léo Belzile"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    lib_dir: "libs"
    chakra: "libs/remark-latest.min.js"
    css: ["default", "css/ath-slides.css", "css/ath-inferno-fonts.css", "css/animate.css"]
    seal: false
    anchor_sections: false
    nature:
      highlightStyle: github
      highlightLines: false
      countIncrementalSlides: false
      ratio: "16:9"
      navigation:
        scroll: false
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, 
                      message = FALSE, 
                      fig.retina = 3, 
                      fig.align = "center",
                      fig.width = 10,
                      fig.asp = 0.618,
                      out.width = "70%")
```
```{r packages-data, echo = FALSE, include=FALSE}
library(knitr)
options(knitr.kable.NA = '')
options(tidyverse.quiet = TRUE)
options(knitr.table.format = "html")
library(tidyverse)
library(patchwork)
```
```{r xaringanExtra, echo=FALSE}
xaringanExtra::use_xaringan_extra(c("tile_view","freezeframe","panelset","clipboard","broadcast"))
```

class: center middle main-title section-title-1

# Linear models

.class-info[

**Session 8**

.light[MATH 80667A: Experimental Design and Statistical Methods <br>for Quantitative Research in Management <br>
HEC Montréal
]

]

---

name: outline
class: title title-inv-1

# Outline
--


.box-4.medium.sp-after-half[Model formulation]

--

.box-5.medium.sp-after-half[Seeking balance]

--

.box-6.medium.sp-after-half[Analysis of trend]
--

.box-7.medium.sp-after-half[Diagnostic plots]

---

layout: false
name: model-formulation
class: center middle section-title section-title-4 animated fadeIn

# Linear model

---


class: title title-4

# Example from the OSC psychology replication

> People can be influenced by the prior consideration of a numerical anchor when forming numerical judgments. [...]  The anchor provides an initial starting point from which estimates are adjusted, and a large body of research demonstrates that adjustment is usually insufficient, leading estimates to be biased towards the initial anchor.

.small[
[Replication of Study 4a of Janiszewski & Uy (2008, Psychological Science) by J. Chandler](https://osf.io/aaudl/)
]

???

People can be influenced by the prior consideration of a numerical anchor when forming numerical judgments. The anchor provides an initial starting point from which estimates are adjusted, and a large body of research demonstrates that adjustment is usually insufficient, leading estimates to be biased towards the initial anchor. Extending this work, Janiszewski and Uy (2008) conceptualized people's attempt to adjust following presentation of an anchor as movement along a subjective representation scale by a certain number of units. Precise numbers (e.g. 9.99) imply a finer-resolution scale than round numbers (e.g. 10). Consequently, adjustment along a subjectively finer resolution scale will result in less objective adjustment than adjustment by the same number of units along a subjectively coarse resolution scale. 

In three experimental studies the authors demonstrate this predicted basic effect and rule out various alternative explanations. Two additional studies (4a and b) found that this effect was especially strong when people were explicitly given more motivation to adjust their estimates (e.g., by implying that the initial anchor substantially overestimated the price). 

---

class: title title-4

# IJLR: It's Just a Linear Regression...

All models we covered so fall ( $t$-tests, one- and two-way ANOVA, factorial designs, latin squares) are all very special
instances of the linear regression model.

The latter says that 
$$\underset{\text{average response}}{\textsf{E}(Y_{i})} = \underset{\text{linear combination of explanatories}}{\beta_0 + \beta_1 \mathrm{X}_{1i} + \cdots + \beta_p \mathrm{X}_{pi}}$$


