---
title: "Repeated measures"
author: "Léo Belzile"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    lib_dir: "libs"
    chakra: "libs/remark-latest.min.js"
    css: ["default", "css/ath-slides.css", "css/ath-inferno-fonts.css", "css/animate.css"]
    seal: false
    anchor_sections: false
    nature:
      highlightStyle: github
      highlightLines: false
      countIncrementalSlides: false
      ratio: "16:9"
      navigation:
        scroll: false
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, 
                      message = FALSE, 
                      fig.retina = 3, 
                      fig.align = "center",
                      fig.width = 10,
                      fig.asp = 0.618,
                      out.width = "70%")
```
```{r packages-data, echo = FALSE, include=FALSE}
library(knitr)
options(knitr.kable.NA = '')
options(tidyverse.quiet = TRUE)
options(knitr.table.format = "html")
library(tidyverse)
library(patchwork)
```
```{r xaringanExtra, echo=FALSE}
xaringanExtra::use_xaringan_extra(c("tile_view","freezeframe","panelset","clipboard","broadcast"))
```

class: center middle main-title section-title-1

# Repeated measures

.class-info[

**Session 10**

.light[MATH 80667A: Experimental Design and Statistical Methods <br>for Quantitative Research in Management <br>
HEC Montréal
]

]

---

name: outline
class: title title-inv-1

# Outline
--


.box-8.large.sp-after-half[Repeated measures]

--

.box-7.large.sp-after-half[Mixed models]
--

.box-6.large.sp-after-half[Testing]
---


layout: false
name: repeated-measures
class: center middle section-title section-title-3

# Repeated measures

